...

### 2023-09-16 manuscript/01_main.Rmd
changed "invertebrate-feeder" to "invertivore"

At the second level, we investigated whether foraging guilds differed in their habitat selection patterns. The original bird observations were summarised by guild so as to obtain for each point abundance values of all the guilds that were observed. Only two guilds had sufficiently high number of overall detections, insectivores (rather, 'invertebrate-feeders') and omnivores. Models were built separately for these two guilds, using the *glmmTMB* package [@brooks2017] to avoid convergence issues. The data summarised at guild level was insufficient for utilisation of two crossed random effects (as was done for the community-level models), so the random effect of sampling day, which had less estimated variance than that of the point, was excluded from the models. This did not result in major differences in the fixed effects estimates and so was not considered to be an issue. The null models had the same fixed effect as in the first case.

Models were constructed following a null model approach, i.e., starting from a null model then building up using stepwise selection of predictors. First, the week variable which had a visible effect (abundances were higher towards the end of the study period; noticed even during the bird counts), and other non-focal variables were added. Following this, the other, focal variables were finally added *(see Table 1)*. Stepwise selection was done by comparing AIC~C~ values of all the models using the AICctab() function from the *bbmle* package [@bolker2020], with an information threshold of 2 dAIC~C~ per df. If a variable was deemed important, its interaction with 'Week' was next tested (i.e., whether or not habitat selection criteria changed with the progress of the season). Once the selection was complete, parsimony was further tested by dropping some complex interaction terms. Finally, the most parsimonious model was selected. At the community level, the final model was a GLMM with Poisson error distribution and log link function fitted with the BOBYQA optimiser *(Table 2)*; the guild-level final models were a Poisson GLMM (log link) for invertebrate-feeders and a negative binomial GLMM (log link) for omnivores.

@@

Within the bird community, habitat selection was very different between invertebrate-feeders and omnivores (Table 3). Invertebrate-feeder abundance was affected by observer, proportion of deciduous trees and dominant ground vegetation. Abundance was lower in sampling points with 'Graminoid' ground vegetation than points with other ground layers (Figure 4). Abundance also decreased with deciduous proportion and, notably, since a polynomial term did not give a better fit, there is no evidence for the notion of mixed forests with intermediate deciduous proportions having higher diversity and abundance. Neither of the two trends changed detectably with time.

@@

The overall abundance of the forest bird community after the breeding season seemed to be higher in patches where the canopy was highly heterogeneous. As invertebrate-feeders did not show any particular relationship with canopy heterogeneity, the trend is probably driven by omnivores. At the beginning of the survey, omnivore abundance too was not related to canopy heterogeneity, but this was completely reversed by the end of the study, with the most heterogeneous points having the highest abundance---twice that of the least heterogeneous points which did not see any change in bird abundance at all over the course of the season. The arrival of migrants may have caused an occupation of the previously 'empty' niche and this was most likely because heterogeneity enables the formation of large foraging flocks, with different individuals in the party focused on different aspects of the habitat.

@@

Birds generally show a hump-shaped trend with proportion of deciduous trees, with highest numbers found in mixed forests and lowest in both purely coniferous and purely broadleaved patches [e.g., @alatalo1981; @jansson2003; @hagar2007; @ohno1997; @felton2010]. However, in alignment with Alatalo's findings [-@alatalo1981] that relative use of such deciduous patches were strong in summer but declined in winter, omnivores in our study in the summer-winter transition showed neither a linear nor a polynomial relationship with proportion of deciduous trees, while invertebrate-feeder abundance declined with deciduous proportion. Oaks are usually linked with higher biodiversity (especially *Q. robur*, which is highly insect-rich and has the highest visitation rates by birds), but oak-rich points in our study were poorest for birds in both periods. It is to be noted that a majority of the herbivorous arthropods on oaks belong to Lepidoptera [45% of the total species richness of phytophages on *Q. robur/petraea*\; @kennedy1984], which attain a peak density in spring. The trees see relatively very low lepidopteran attacks after mid-June [@feeny1970]. When considering all phytophagous insects, there is a second but much smaller peak in abundance in late summer and early autumn of suckers, leaf-miners and gall-formers [@southwood2004]. Thus, the overwhelming preference for oaks by arthropods, and consequently by predatory birds, is likely to be prevalent mainly in spring and summer. The arthropod communities of oaks and their regulation by predators has rarely been explored in the post-breeding periods of late summer and early autumn. Moreover, the specific invertebrate-feeding species present during this summer-autumn transition period such as treecreepers, robin and goldcrest/firecrest are largely specialists and do not prey on external-feeding herbivores. Treecreepers are bark-feeders, robin is mostly a ground-forager and goldcrest is specialised for spruce [firecrest shows no preference for either conifers or broadleaf trees\; @leisler1982].

@@

The general abundance increase in the bird community as the season progressed was likely not due to better detection rates as the birds became bolder and more active, but rather due to migrants arriving from the north to winter here, although trapping and tracking of individual birds is necessary to confirm this. Further, while abundance of omnivores increased with time, that of invertebrate-feeders remained largely constant. This agrees with the fact that most of the northern migrants that arrive here are omnivores (tits), also suggesting that detectability of the naturally quiet invertebrate-feeders was not a major issue in this study. Unexpectedly, there were very few instances of birds feeding on fruits. This could be either because arthropod abundance in the forest was still high enough to support the birds, or because the birds in this region---mostly non-migrants or partial omnivorous migrants---simply do not shift their foraging strategy, unlike birds elsewhere [see @parrish1997; @parrish2000], and rather only feed on fruits facultatively.

@@

The two guilds showing different relationships with habitat characteristics is indirect evidence of differences in their resource-use strategies. Invertebrate-feeders were selective only about one major characteristic, but were highly specific and consistent about it: they avoided oak-dominated homogeneous patches of forest, preferring spruce patches with heterogeneity, and were otherwise broad in resource-use. Omnivores, on the other hand, were affected by a larger interplay of factors and their patterns changed over time, probably because they had to adjust and accommodate the new arrivals. They did not seem to be specific about food resources directly. Rather, they showed a slight structural preference for forest edges with low densities of trees and this remained constant with time, but they formed an additional strong preference for highly heterogeneous areas as the season progressed. The latter could have been extremely important in accommodating the northern arrivals and presumably also enabled the large mixed-species foraging flocks. @alatalo1981 found evidence for the prevalence of interspecific competition in the seasonal environment of Finland. In our case, such competition brought about by migrant arrival was presumably mitigated by the omnivores choosing to form mixed-species flocks. The model for omnivores had a negative binomial distribution, implying overdispersion in omnivore abundance values, corroborating the presence of such flocks.

@@

***Table 3: GLMMs of guild abundances with habitat variables. Model names 'inv.x' refer to invertebrate-feeder models while names 'omn.x' refer to omnivore models. Models at each step of term addition/deletion are shown with their corresponding information such as degrees of freedom (df), predictor terms, log-likelihood, AICC (which was used as a selection criterion) and Akaike weight. 'sTPD': scaled proportion of deciduous trees; 'DOM': dominant ground vegetation; 'HC': habitat class; 'CH': canopy heterogeneity; 'TD': tree density.***

>>>>

At the second level, we investigated whether foraging guilds differed in their habitat selection patterns. The original bird observations were summarised by guild so as to obtain for each point abundance values of all the guilds that were observed. Only two guilds had sufficiently high number of overall detections, insectivores (rather, 'invertivores') and omnivores. Models were built separately for these two guilds, using the *glmmTMB* package [@brooks2017] to avoid convergence issues. The data summarised at guild level was insufficient for utilisation of two crossed random effects (as was done for the community-level models), so the random effect of sampling day, which had less estimated variance than that of the point, was excluded from the models. This did not result in major differences in the fixed effects estimates and so was not considered to be an issue. The null models had the same fixed effect as in the first case.

Models were constructed following a null model approach, i.e., starting from a null model then building up using stepwise selection of predictors. First, the week variable which had a visible effect (abundances were higher towards the end of the study period; noticed even during the bird counts), and other non-focal variables were added. Following this, the other, focal variables were finally added *(see Table 1)*. Stepwise selection was done by comparing AIC~C~ values of all the models using the AICctab() function from the *bbmle* package [@bolker2020], with an information threshold of 2 dAIC~C~ per df. If a variable was deemed important, its interaction with 'Week' was next tested (i.e., whether or not habitat selection criteria changed with the progress of the season). Once the selection was complete, parsimony was further tested by dropping some complex interaction terms. Finally, the most parsimonious model was selected. At the community level, the final model was a GLMM with Poisson error distribution and log link function fitted with the BOBYQA optimiser *(Table 2)*; the guild-level final models were a Poisson GLMM (log link) for invertivores and a negative binomial GLMM (log link) for omnivores.

@@

Within the bird community, habitat selection was very different between invertivores and omnivores (Table 3). Invertivore abundance was affected by observer, proportion of deciduous trees and dominant ground vegetation. Abundance was lower in sampling points with 'Graminoid' ground vegetation than points with other ground layers (Figure 4). Abundance also decreased with deciduous proportion and, notably, since a polynomial term did not give a better fit, there is no evidence for the notion of mixed forests with intermediate deciduous proportions having higher diversity and abundance. Neither of the two trends changed detectably with time.

@@

The overall abundance of the forest bird community after the breeding season seemed to be higher in patches where the canopy was highly heterogeneous. As invertivores did not show any particular relationship with canopy heterogeneity, the trend is probably driven by omnivores. At the beginning of the survey, omnivore abundance too was not related to canopy heterogeneity, but this was completely reversed by the end of the study, with the most heterogeneous points having the highest abundance---twice that of the least heterogeneous points which did not see any change in bird abundance at all over the course of the season. The arrival of migrants may have caused an occupation of the previously 'empty' niche and this was most likely because heterogeneity enables the formation of large foraging flocks, with different individuals in the party focused on different aspects of the habitat.

@@

Birds generally show a hump-shaped trend with proportion of deciduous trees, with highest numbers found in mixed forests and lowest in both purely coniferous and purely broadleaved patches [e.g., @alatalo1981; @jansson2003; @hagar2007; @ohno1997; @felton2010]. However, in alignment with Alatalo's findings [-@alatalo1981] that relative use of such deciduous patches were strong in summer but declined in winter, omnivores in our study in the summer-winter transition showed neither a linear nor a polynomial relationship with proportion of deciduous trees, while invertivore abundance declined with deciduous proportion. Oaks are usually linked with higher biodiversity (especially *Q. robur*, which is highly insect-rich and has the highest visitation rates by birds), but oak-rich points in our study were poorest for birds in both periods. It is to be noted that a majority of the herbivorous arthropods on oaks belong to Lepidoptera [45% of the total species richness of phytophages on *Q. robur/petraea*\; @kennedy1984], which attain a peak density in spring. The trees see relatively very low lepidopteran attacks after mid-June [@feeny1970]. When considering all phytophagous insects, there is a second but much smaller peak in abundance in late summer and early autumn of suckers, leaf-miners and gall-formers [@southwood2004]. Thus, the overwhelming preference for oaks by arthropods, and consequently by predatory birds, is likely to be prevalent mainly in spring and summer. The arthropod communities of oaks and their regulation by predators has rarely been explored in the post-breeding periods of late summer and early autumn. Moreover, the specific invertebrate-feeding species present during this summer-autumn transition period such as treecreepers, robin and goldcrest/firecrest are largely specialists and do not prey on external-feeding herbivores. Treecreepers are bark-feeders, robin is mostly a ground-forager and goldcrest is specialised for spruce [firecrest shows no preference for either conifers or broadleaf trees\; @leisler1982].

@@

The general abundance increase in the bird community as the season progressed was likely not due to better detection rates as the birds became bolder and more active, but rather due to migrants arriving from the north to winter here, although trapping and tracking of individual birds is necessary to confirm this. Further, while abundance of omnivores increased with time, that of invertivores remained largely constant. This agrees with the fact that most of the northern migrants that arrive here are omnivores (tits), also suggesting that detectability of the naturally quiet invertivores was not a major issue in this study. Unexpectedly, there were very few instances of birds feeding on fruits. This could be either because arthropod abundance in the forest was still high enough to support the birds, or because the birds in this region---mostly non-migrants or partial omnivorous migrants---simply do not shift their foraging strategy, unlike birds elsewhere [see @parrish1997; @parrish2000], and rather only feed on fruits facultatively.

@@

The two guilds showing different relationships with habitat characteristics is indirect evidence of differences in their resource-use strategies. Invertivores were selective only about one major characteristic, but were highly specific and consistent about it: they avoided oak-dominated homogeneous patches of forest, preferring spruce patches with heterogeneity, and were otherwise broad in resource-use. Omnivores, on the other hand, were affected by a larger interplay of factors and their patterns changed over time, probably because they had to adjust and accommodate the new arrivals. They did not seem to be specific about food resources directly. Rather, they showed a slight structural preference for forest edges with low densities of trees and this remained constant with time, but they formed an additional strong preference for highly heterogeneous areas as the season progressed. The latter could have been extremely important in accommodating the northern arrivals and presumably also enabled the large mixed-species foraging flocks. @alatalo1981 found evidence for the prevalence of interspecific competition in the seasonal environment of Finland. In our case, such competition brought about by migrant arrival was presumably mitigated by the omnivores choosing to form mixed-species flocks. The model for omnivores had a negative binomial distribution, implying overdispersion in omnivore abundance values, corroborating the presence of such flocks.

@@

***Table 3: GLMMs of guild abundances with habitat variables. Model names 'inv.x' refer to invertivore models while names 'omn.x' refer to omnivore models. Models at each step of term addition/deletion are shown with their corresponding information such as degrees of freedom (df), predictor terms, log-likelihood, AICC (which was used as a selection criterion) and Akaike weight. 'sTPD': scaled proportion of deciduous trees; 'DOM': dominant ground vegetation; 'HC': habitat class; 'CH': canopy heterogeneity; 'TD': tree density.***
